---
title: "PSTAT HW 4"
author: "TJ Sipin"
date: "3/13/2022"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, cache = T, message = F)
library(tidyverse)
library(ROCR)
library(ggridges)
library(dendextend)
library(tidymodels)


knitr::opts_chunk$set(echo = TRUE)


leukemia_data <- read_csv("leukemia_data.csv")
```

## Clustering and dimension reduction for gene expression data

(a) The class of the first column of leukemia_data, Type, is set to character by default. (1 pts) Convert the Type column to a factor using the mutate function. (2 pts) Use the table command to print the number of patients with each leukemia subtype. (1 pts) Which leukemia subtype occurs the least in this data?

```{r}
# Converting Type to factor
leukemia_data <- leukemia_data %>%
  mutate(Type = as.factor(Type))

# Table of number of patients with each leukemia subtype
leukemiaSubtype <- table(leukemia_data[ , "Type"])
leukemiaSubtype

# Subtype with lowest number of occurrences
print(paste("Subtype with lowest number of occurrences: ",names(which.min(leukemiaSubtype))))
```
(b) (2 pts) Run PCA on the leukemia data using prcomp function with scale=TRUE and center=TRUE (this scales each gene to have mean 0 and variance 1). Make sure you exclude the Type column when you run the PCA function (we are only interested in the unsupervised learning regime for now, where our focus is on reducing the dimension of the gene expression values). (2 pts) Plot the proportion of variance explained by each principal component (PVE) and the cumulative PVE.
(2 pts) How many PCs do we need in order to explain 90% of the total variation in the data?)
```{r}
# Recipe
recipe <- recipe(Type ~ ., data = leukemia_data)
```

```{r, results="hide", warning=FALSE, message=FALSE}
# Perform PCA on leukemia_data
pr.out = prcomp(leukemia_data[2:ncol(leukemia_data)], scale = T, center = T)

# Plotting proportion of variance explained by each PC (PVE) and the cumulative PVE

pca_trans <- recipe %>%
  step_center(all_numeric()) %>%
  step_scale(all_numeric()) %>%
  step_pca(all_numeric())

pca_estimates <- prep(pca_trans)

sdev <- pca_estimates$steps[[3]]$res$sdev

percent_variation <- sdev^2 / sum(sdev^2)

var_df <- data.frame(PC=paste0("PC", 1:length(sdev)),
                     var_explained=percent_variation,
                     stringsAsFactors = F)

var_df %>%
  mutate(PC = fct_inorder(PC)) %>%
  ggplot(aes(x=PC, y=var_explained)) +
  geom_col()

plot(cumsum(percent_variation), xlab = "Principal Component",
     ylab = "Proportion of Variation Explained",
     ylim = c(0,1), type = 'b') %>%
  abline(h=.9, col = 2) %>%
  grid(col = "lightgray", lty = "dotted")
```




It appears that 200 PCs are needed to explain 90% of the total variation in the data.


(c) (2 pts) Use the results of PCA to project the data into the first two principal component dimensions, i.e., compute the scores vectors for the first two PCs. Recall in Lab 9, prcomp returns this dimension reduced data in the columns of x. Plot the data as a scatter plot using plot function with col=plot_colors. (2 pts) Add the leukemia type labels to the plot using text with labels argument set to the leukemia type and the col to plot_colors (it may help legibility to make the points on the plot very small by setting cex to a small number). (2 pts) Which group is most clearly separated from the others along the PC2 axis? (2 pts) Which genes have the highest absolute loadings for PC1 (the genes that have the largest (absolute) weights in the weighted average used to create the new variable PC1)? You can find these by taking the absolute values of the first principal component loadings and sorting them. Print the first 6 genes in this sorted vector using the head function.

```{r}
rainbow_colors <- rainbow(7)
plot_colors = rainbow_colors[(leukemia_data$Type)]

# Using lab 09
pr.out = prcomp(leukemia_data[2:ncol(leukemia_data)], scale = T, center = T)
pr.out$x[,1:2]

plot(x = pr.out$x[,1:2][,1], 
     y = pr.out$x[,1:2][,2],
     col = plot_colors)+
  text(pr.out$x[,1:2], labels = leukemia_data$Type, cex = 0.3)

```

Along the PC2 axis, it seems like E2A-PBX1 or T-ALL is the most clearly separated.

```{r}
head(sort(abs(pr.out$rotation[,1]), decreasing = T), 6)
```

The above show the six genes with the highest loading in PC1.

(d) PCA orders the principal components according to the amount of total variation in the data that they explain. This does not mean, however, that the principal components are sorted in terms of how useful they are at capturing variation between the leukemia groups. For example, if gene expression varied significantly with age and gender (independent of leukemia status), the first principal components could reflect genetic variation due to age and gender, but not to leukemia. (4 pts) See if the 3rd PC is better at discriminating between leukemia types by plotting the data projected onto the first and third principal components (not the second).

```{r}
plot(x = pr.out$x[, c(1, 3)][,1], 
     y = pr.out$x[, c(1,3)][,2],
     col = plot_colors)+
  text(pr.out$x[, c(1,3)], labels = leukemia_data$Type, cex = 0.3)
```

There seems to be much better separation (discrimination) between leukemia types as the variance is high but the spread of the means are low.

(e) (2 pts) Use the filter command to create a new dataframe (or a tibble if you’d like) leukemia_subset by subsetting to include only rows for which Type is either T-ALL, TEL-AML1, or Hyperdip50. (2 pts) Compute a euclidean distance matrix between the subjects using the dist function and then run hierarchical clustering using complete linkage. (2 pts) Plot two dendrograms based on the hierarchical clustering result. In the first plot, force 3 leukemia types to be the labels of terminal nodes, color the branches and labels to have 3 groups and rotate the dendrogram counter-clockwise to have all the terminal nodes on the right. (2 pts) In the second plot, do all the same things except that this time color all the branches and labels to have 5 groups. Please make sure library dendextend is installed. (See Lab 8)

```{r, results='hide'}
leukemia_subset <- filter(leukemia_data, Type == "T-ALL" | Type == "TEL-AML1" | Type == "Hyperdip50")
leukemia_subset
```


First plot colored by three clusters:

```{r}
leukemia.dist <- dist(leukemia_subset)
leukemia.hclust <- hclust(leukemia.dist)
dend3 = as.dendrogram(leukemia.hclust)
dend3 = color_branches(dend3, k=3)
dend3 = color_labels(dend3, k=3)

dend3 = set(dend3, "labels_cex", 0.1)

dend3 = set_labels(dend3, labels=leukemia_subset$Type[order.dendrogram(dend3)])

# Plot dendogram
plot(dend3, horiz=T, main = "Dendogram colored by three clusters")
```

Second plot with 5 groups:

```{r}
leukemia.dist <- dist(leukemia_subset)
leukemia.hclust <- hclust(leukemia.dist)
dend5 = as.dendrogram(leukemia.hclust)
dend5 = color_branches(dend5, k=5)
dend5 = color_labels(dend5, k=5)

dend5 = set(dend5, "labels_cex", 0.1)

dend5 = set_labels(dend5, labels=leukemia_subset$Type[order.dendrogram(dend3)])

# Plot dendogram
plot(dend5, horiz=T, main = "Dendogram colored by three clusters")
```




